Estimating the distribution of block sizes as a function of average transaction rates

### 2015-10-22 paper/paper.tex
Ran a spellcheck and made minor changes

There is always some chance that a given block will fill up since arrival times are stochastic. Specifically, for the $k$'th block this probability is
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right).
\]
This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is determinstic with a constant transaction rate of $\gamma$ MB/min, then 
\[
x_{\tau_k} - x_{\tau_{k-1}} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
\]
and the probability of a full block is available in closed form as
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
\]

@@

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider nonnegative infinite activity L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$. The following discussion is brief -- see e.g. \citet{barndorff2012basics} for a fuller treatment. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. The gamma process is called an infinite activitive process since for any time $\tau$ and increment $\delta>0$, $x_{\tau + \delta} - x_{\tau} > 0$.

@@

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the conditional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar distribution that avoids numerical problems is the lognormal, which we will use.

@@

vvvvvvvvvvvvvvvvvvvv
\section{Discussion}
Our analysis does not resolve the contentious block size issue, but it is an important input into weighing the tradeoff between supporting higher transaction rates and the security issues created by larger blocks. For example, at the existing 1 MB maximum, as average transaction rates increase from 500 KB/block to 600 KB/block, an additional five percent or so of blocks will hit the limit. These full blocks signal that not all transactions are being included in the first block after their transmission.

In addition, a greater understanding of the dynamics of full blocks versus the average transaction rates could be used to help clients set fees appropriately, as a function of the observed rate of transactions over the last several blocks and the existing maximum block size. Creating a more dynamic fee market is another way, besides simply increasing the maximum block size, to scale the network effectively.
^^^^^^^^^^^^^^^^^^^^

>>>>

There is always some chance that a given block will fill up since arrival times are stochastic. Specifically, for the $k$'th block this probability is
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right).
\]
This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is deterministic with a constant transaction rate of $\gamma$ MB/min, then 
\[
x_{\tau_k} - x_{\tau_{k-1}} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
\]
and the probability of a full block is available in closed form as
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
\]

@@

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official time-stamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' time-stamps, occasionally a block has a time-stamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid time-stamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider nonnegative infinite activity L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$. The following discussion is brief -- see e.g. \citet{barndorff2012basics} for a fuller treatment. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continuous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. The gamma process is called an infinite activity process since for any time $\tau$ and increment $\delta>0$, $x_{\tau + \delta} - x_{\tau} > 0$.

@@

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the conditional distribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar distribution that avoids numerical problems is the lognormal, which we will use.

@@

\section{Discussion}
Our analysis does not resolve the contentious block size issue, but it is an important input into weighing the tradeoff between supporting higher transaction rates and the security issues created by larger blocks. For example, at the existing 1 MB maximum, as average transaction rates increase from 500 KB/block to 600 KB/block, an additional five percent or so of blocks will hit the limit. These full blocks signal that not all transactions are being included in the first block after their transmission. In addition, a greater understanding of the dynamics of full blocks versus the average transaction rates could be used to help clients set fees appropriately, as a function of the observed rate of transactions over the last several blocks and the existing maximum block size. Creating a more dynamic fee market is another way, besides simply increasing the maximum block size, to scale the network effectively.
